Solving the Nonstationary Richards Equation\\ With Adaptive $hp$-FEM

### 2011-03-14 paper.tex
added requested chapters

@@

@@

%odkomentoval jsem, protoze se na tuto analytickou rovnici budu dal v textu odkazovat
When van Genuchten's formula (\ref{vangenuchten-eq}) is substitutedin to this integral 
law, the following analytical formula is obtained

  \begin{equation}
  \label{mualem-anal}
  K(h) = \left\{
    \begin{array}{l l}
    K_s \frac{\left( 1- (-\alpha h)^{nm} \left( 1+ (-\alpha h)^n \right)^{-m} \right)^2}{\left(1+(-\alpha h)^n \right)^{\frac{m}{2}}} \\
    \mbox{for all $h \in$ $(-\infty,0)$};\\
    K_s  \ \mbox{for all $h \in$ $\langle 0, +\infty)$}\\
    \end{array} \right.
    \end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Since the treatment of the initial condition (\ref{ic}) is nontrivial
in the finite element context, let us say a few words about it. 
At time $t = 0$ the exact solution (\ref{ic}) is constant and it 
does not satisfy the Dirichlet boundary condition on the top edge. 
Therefore we begin at a small positive time $t = 10^{-3}$.
Starting from a coarse low-order mesh, automatic adaptivity is used
to obtain a piecewise-polynomial approximate initial condition 
$h_0$ that approximates the (exact) initial condition $h$ with 
a sufficiently small relative error $TOL$ in the $H^1$-norm. The 
spatial accuracy for all subsequent computations is set to $TOL = 1\%$.
The function $h_0$ is shown in Fig. \ref{fig:init-cond}.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[height=4.5cm]{mesh_init_h1.png}
\end{center}
\vspace{-6mm}
\caption{Initial mesh for $h$-FEM with linear elements (24843 DOF).}
\label{fig:init-meshes-3}
\end{figure}
In this and all consequent examples we use the implicit second-order 
Crank-Nicolson method for time integration.
In Figs. \ref{fig:init-meshes-1} - \ref{fig:init-meshes-3} 
different colors are used to distinguish between 
different polynomial degrees of finite elements. All elements in these computations 
are quadrilaterals. Notice in Fig. \ref{fig:init-meshes-1} that adaptive $hp$-FEM employs
anisotropic (i.e., directionally different) polynomial degrees in finite 
elements -- this is when an element in the mesh contains two different colors. 
The steady-state solution is shown in Fig. \ref{fig:steady-state}.

@@

Due to a large amount of computer operations involved in the evaluation of both 
the van Genuchten's model~(\ref{vangenuchten-eq}) and Mualem's model~(\ref{mualem-anal}), 
the functions $C(h)$ and $K(h)$ were replaced with quintic splines. Also the first derivative
of $C(h)$ and the first and second derivatives of $K(h)$ were approximated in this way 
for the Newton's method.
Via this approximation we cured the problem of singularity of the second derivative 
of $K(h)$,

$$
  \lim_{h \to 0-} K''(h) = -\infty \quad \mbox{but} \quad \lim_{h \to 0+} K''(h) = 0. 
$$ 
The spline approximation $p(h)$ has 
$$
  p''(h=0) = 0,
$$
as shown in Fig. \ref{fig:kh-approx} (for gravel).
This spline only differs from the original function $K(h)$ in the first interval 
$(-1, 0)$ and in the remaining intervals it is virtually identical to the 
original function. Close to full saturation, the approximating curve's shape is 
very similar to the modified van Genuchten Mualem model that was recommended by Vogel et al. 
in~\cite{vogel}. Compared to the original Vogel's approach, our spline approximation 
preserves the smoothness of the $K(h)$ function in its entire domain of definition.
Unsaturated hydraulic properties of all materials are summarized in Table \ref{tab:mat-example3}.

@@

To be done by Michal.\\

Zde mame strucne popsat co jsme udelali, a rict ze
Crank-Nicolson neni nase finalni volba a ze se 
chystame zlepsit casovou integraci. Mame dat reference
na Keese a Williamse, viz. soubor review\_kees.txt.

Je treba sem napsat ze umoznime v Hermesu casovou integraci 
zalozenou na libovolnych Butcherovych tabulkach,
a ze potom bude adaptivita cas. kroku prirozene 
mozna na zaklade vnorenych (embedded) R-K methods.

@@

To be done by Michal.\\

Kees rikal ze toto je hodne dulezite a musi se to udelat poradne.

>>>>

\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}%[dvips]
\usepackage{amsmath}
\usepackage{latexsym}
\usepackage{pstricks}
\usepackage{mathrsfs}

@@

%%%%%%%%%%%%%new commands%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{definition}[1][Definition]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}
\newenvironment{example}[1][Example]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\newcommand{\icom}{{\mbox{\rm $\scriptscriptstyle ^\mid$\hspace{-0.40em}C}}}
\newcommand{\om}{\Omega}
\newcommand{\pom}{\partial\Omega}
\newcommand{\epal}{{\mathcal E}(\overline{\om})}
\newcommand{\ak}{A_k}
\newcommand{\bk}{B_k}
\newcommand{\buuv}{b(\bfu,\bfu,\bfv)}
\newcommand{\buwt}{b(\bfu,\bfw,.)}
\newcommand{\buwv}{b(\bfu,\bfw,\bfv)}
\newcommand{\bwuv}{b(\bfw,\bfu,\bfv)}
\newcommand{\cco}{c = c(\om)}
\newcommand{\ccot}{c = c(\om,T)}
\newcommand{\cd}{c_2}
\newcommand{\cdco}{c_2 = c_2(\om)}
\newcommand{\cj}{c_1}
\newcommand{\cjco}{c_1 = c_1(\om)}
\newcommand{\cjcot}{c_1 = c_1(\om,T)}
\newcommand{\ct}{c_3}
\newcommand{\ctco}{c_3 = c_3(\om)}
\newcommand{\cy}{c_4}
\newcommand{\cyco}{c_4 = c_4(\om)}
\newcommand{\ddt}{{{d}\over{dt}}}
\newcommand{\dom}{\mbox{ }d(\om)}
\newcommand{\ds}{\mbox{ }ds}
\newcommand{\dt}{\mbox{ }dt}
\newcommand{\dpg}{\mbox{ }d(\partial\om)}
\newcommand{\dc}{\|}
\newcommand{\dz}{\bigl(\!\bigl(}
\newcommand{\bdz}{(\!(}
\newcommand{\dzr}{)\!)}
\newcommand{\eij}{e_{ij}}
\newcommand{\eiju}{e_{ij}(\bfu)}
\newcommand{\eijv}{e_{ij}(\bfv)}
\newcommand{\eijw}{e_{ij}(\bfw)}
\newcommand{\eu}{{\mathcal A}_u}
\newcommand{\fk}{f_k}
\newcommand{\fn}{f_0}
\newcommand{\gj}{\Gamma_1}
\newcommand{\gd}{\Gamma_2}
\newcommand{\gh}{{\mathcal G}_H}
\newcommand{\gv}{{\mathcal G}}
\newcommand{\gt}{\Gamma_3}
\newcommand{\gc}{\Gamma_4}
\newcommand{\ha}{H}
\newcommand{\haa}{H}
\newcommand{\intnt}{\int_0^T}
\newcommand{\intnst}{\int_0^{T^*}}
\newcommand{\iom}{\int_\om}
\newcommand{\isom}{\int_{\partial\om}}
\newcommand{\kc}{K_C}
\newcommand{\pcx}{P_c}
\newcommand{\ldd}{L^2(0,T;\,D)}
\newcommand{\ldsd}{L^2(0,\ts;\,D)}
\newcommand{\ldh}{L^2(0,T;\,H)}
\newcommand{\ldsh}{L^2(0,\ts;\,H)}
\newcommand{\ldon}{L^2(\om)^m}
\newcommand{\ldv}{L^2(0,T;\,\vajd)}
\newcommand{\ldsv}{L^2(0,\ts;\,\vajd)}
\newcommand{\ldvso}{L^2(0,T;\,\vso)}
\newcommand{\lnvso}{L^{\infty}(0,T;\,\vso)}
\newcommand{\lnv}{L^{\infty}(0,T;\,\vajd)}
\newcommand{\lnsv}{L^{\infty}(0,\ts;\,\vajd)}
\newcommand{\lnvh}{L^{\infty}(0,T;\,\vajdh)}
\newcommand{\ldvh}{L^2(0,T;\,\vajdh)}
\newcommand{\lnh}{L^\infty(0,T;\,H)}
\newcommand{\lno}{L^\infty(\om)^3}
\newcommand{\lco}{L^4(\om)^m}
\newcommand{\ldor}{L^2(\om)/\iree}
\newcommand{\mc}{{\mathcal M}_C}
\newcommand{\mn}{M_0}
\newcommand{\mr}{{\mathcal M}_R}
\newcommand{\nd}{\|_D}
\newcommand{\njd}{\|_{V^{1,2}}}
\newcommand{\nnd}{\|_{H}}
\newcommand{\njp}{\|_{V^{1,p}}}
\newcommand{\njq}{\|_{V^{1,q}}}
\newcommand{\njqh}{\|_{({V^{1,q}})^*}}
\newcommand{\nldd}{\|_{L^2(0,T;\,D)}}
\newcommand{\nldv}{\|_{\ldv}}
\newcommand{\nldvh}{\|_{L^2(0,T;\,\vajd^*)}}
\newcommand{\nlnv}{\|_{\lnv}}
\newcommand{\nlnvso}{\|_{L^\infty(0,T;\,\vajd^{{7/8},2})}}
\newcommand{\nldvso}{\|_{L^2(0,T;\,\vajd^{{7/8},2})}}
\newcommand{\nlnvh}{\|_{\lnvh}}
\newcommand{\nldh}{\|_{L^2(0,T;\,H)}}
\newcommand{\nlnh}{\|_{L^\infty(0,T;\,\haa)}}
\newcommand{\nha}{\|_{\haa}}
\newcommand{\nlco}{\|_{\lco}}
\newcommand{\nre}{\|_E}
\newcommand{\nt}{(0,T)}
\newcommand{\nv}{\|_V}
\newcommand{\nvjdh}{\|_{\vajdh}}
\newcommand{\nvso}{\|_{\vso}}
\newcommand{\nvrd}{\|_{\vrd}}
\newcommand{\nvnth}{\|_{\vnth}}
\newcommand{\nwjd}{\|_{\wjdo}}
\newcommand{\nwkd}{\|_{\wkd}}
\newcommand{\nx}{\|_X}
\newcommand{\nxj}{\|_{\xj}}
\newcommand{\nxd}{\|_{\xd}}
\newcommand{\nxp}{\|_{\xp}}
\newcommand{\ny}{\|_{X_2}}
\newcommand{\nyp}{\|_{\yp}}
\newcommand{\ob}{{\mathcal B}}
\newcommand{\obp}{{\mathcal B}_{\mathcal P}}
\newcommand{\oc}{{\mathcal C}}
\newcommand{\ocp}{{\mathcal C}_{\mathcal P}}
\newcommand{\ogu}{{\mathcal G}_u}
\newcommand{\ogpu}{{{\mathcal G}_{\mathcal P}}_{\bfu}}
\newcommand{\oh}{{\cal H}}
\newcommand{\ohj}{{\cal H}_1}
\newcommand{\ohd}{{\cal H}_2}
\newcommand{\oht}{{\cal H}_3}
\newcommand{\okc}{\overline{\kc}}
\newcommand{\oku}{{\cal K}_u}
\newcommand{\okpu}{{{\cal K}_{\cal P}}_{\bfu}}
\newcommand{\omnt}{\om\times\nt}
\newcommand{\omu}{\bfgamma_0}
\newcommand{\on}{{\cal N}}
\newcommand{\onp}{{\cal N}_{\cal P}}
\newcommand{\pres}{{\cal P}}
\newcommand{\pu}{P;\bfu}
\newcommand{\phik}{\bfPhi_k}
\newcommand{\psik}{\bfPsi_k}
\newcommand{\sh}{\bigr)\!\bigr)_H}
\newcommand{\bsh}{\Bigr)\!\Bigr)_H}
\newcommand{\bshaa}{\bigr)\!\bigr)_{0,2}}
\newcommand{\swnjdo}{)\!)_{\wnjdo}}
\newcommand{\sd}{)\!)_D}
\newcommand{\svjd}{)\!)_V}
\newcommand{\svajd}{\bigr)\!\bigr)_V}
\newcommand{\bsvjd}{\bigr)\!\bigr)_V}
\newcommand{\bsvajd}{\bigr)\!\bigr)_{1,2}}
\newcommand{\sldv}{)\!)_{\ldv}}
\newcommand{\sumkjinf}{\sum_{k=1}^{\infty}}
\newcommand{\os}{{\cal S}}
\newcommand{\osp}{{\cal S}_{\cal P}}
\newcommand{\tc}{\|}
\newcommand{\ts}{T^*}
\newcommand{\tnt}{t\in(0,T)}
\newcommand{\tnts}{t\in(0,\ts)}
\newcommand{\ud}{\bfu_B}
\newcommand{\uj}{{\bfu}_A}
\newcommand{\uk}{\bfu_k}
\newcommand{\unu}{{\bfu}_0}
\newcommand{\vjd}{V}
\newcommand{\vajd}{V}
\newcommand{\vakp}{V^{k,p}}
\newcommand{\vajdh}{\vajd^*}
\newcommand{\vso}{V_\kappa^{{7/8},2}}
\newcommand{\vrd}{V^{r_0,2}}
\newcommand{\vvajd}{\bfv\in\vajd}
\newcommand{\vnq}{V^{0,q}}
\newcommand{\vnt}{V^{0,3}}
\newcommand{\vnth}{(V^{0,3})^*}
\newcommand{\wjdo}{W^{1,2}(\om)^m}
\newcommand{\wnjdo}{W_0^{1,2}(\om)}
\newcommand{\wnjdoh}{W_0^{1,2}(\om)^*}
\newcommand{\wnjdon}{W_0^{1,2}(\om)^m}
\newcommand{\wnjdonh}{(W_0^{1,2}(\om)^m)^*}
\newcommand{\wjdp}{W^{1,{12/5}}(\om)}
\newcommand{\wk}{\bfw_k}
\newcommand{\wkd}{W^{k,2}(\om)^m}
\newcommand{\wkci}{\bfw_{k,i}}
\newcommand{\wkcj}{\bfw_{k,j}}
\newcommand{\wki}{\bfw_{k_i}}
\newcommand{\wsi}{{\bf w^*_i}}
\newcommand{\wsj}{{\bf w^*_j}}
\newcommand{\nwjdp}{\|_{\wjdp}}
\newcommand{\xp}{X_{\cal P}}
\newcommand{\xv}{X}
\newcommand{\xj}{X_{1,\ts}}
\newcommand{\xd}{X_{2,\ts}}
\newcommand{\yp}{X_2_{\cal P}}
\newcommand{\yv}{X_2}
\newcommand{\zrx}{\bfzero_X}
\newcommand{\zry}{\bfzero_{X_2}}
\newcommand{\zrxp}{\bfzero_{\xp}}
\newcommand{\zryp}{\bfzero_{\yp}}
\newcommand{\bfb}{\mbox{\boldmath{$b$}}}
% \newcommand{\bfx}{\mbox{\boldmath{$x$}}}
% \newcommand{\bfn}{\mbox{\boldmath{$n$}}}
\newcommand{\bfe}{\mbox{\boldmath{$e$}}}
\newcommand{\bff}{\mbox{\boldmath{$f$}}}
\newcommand{\bfF}{\mbox{\boldmath{$F$}}}
\newcommand{\bfg}{\mbox{\boldmath{$g$}}}
\newcommand{\bfq}{\mbox{\boldmath{$q$}}}
\newcommand{\bfh}{\mbox{\boldmath{$h$}}}
% \newcommand{\bfu}{\mbox{\boldmath{$u$}}}
\newcommand{\bfw}{\mbox{\boldmath{$w$}}}
% \newcommand{\bfv}{\mbox{\boldmath{$v$}}}
% \newcommand{\bfz}{\mbox{\boldmath{$z$}}}
\newcommand{\bfL}{\mbox{\boldmath{$L$}}}
\newcommand{\bfa}{\mbox{\boldmath{$a$}}}
% \newcommand{\bfU}{\mbox{\boldmath{$U$}}}
% \newcommand{\bfM}{\mbox{\boldmath{$M$}}}
\newcommand{\bfN}{\mbox{\boldmath{$N$}}}
\newcommand{\bfE}{\mbox{\boldmath{$E$}}}
% \newcommand{\bfC}{\mbox{\boldmath{$C$}}}
% \newcommand{\bfS}{\mbox{\boldmath{$S$}}}
\newcommand{\bfJ}{\mbox{\boldmath{$J$}}}
% \newcommand{\bfy}{\mbox{\boldmath{$y$}}}
\newcommand{\bfp}{\mbox{\boldmath{$p$}}}
\newcommand{\bfK}{\mbox{\boldmath{$K$}}}
\newcommand{\bfI}{\mbox{\boldmath{$I$}}}
\newcommand{\bfP}{\mbox{\boldmath{$P$}}}
\newcommand{\bfB}{\mbox{\boldmath{$B$}}}
\newcommand{\bfH}{\mbox{\boldmath{$H$}}}
\newcommand{\bfV}{\mbox{\boldmath{$V$}}}
\newcommand{\bfW}{\mbox{\boldmath{$W$}}}
% \newcommand{\bfD}{\mbox{\boldmath{$D$}}}
% \newcommand{\bfR}{\mbox{\boldmath{$R$}}}
\newcommand{\bfmathcalF}{\mbox{\boldmath{$\mathcal{F}$}}}
\newcommand{\bfmathcalA}{\mbox{\boldmath{$\mathcal{A}$}}}
% \newcommand{\bfX}{\mbox{\boldmath{$X$}}}
\newcommand{\bfmu}{\mbox{\boldmath{$\mu$}}}
\newcommand{\bflambda}{\mbox{\boldmath{$\lambda$}}}
\newcommand{\bfomega}{\mbox{\boldmath{$\omega$}}}
% \newcommand{\bfphi}{\mbox{\boldmath{$\phi$}}}
\newcommand{\bfeta}{\mbox{\boldmath{$\eta$}}}
\newcommand{\bftau}{\mbox{\boldmath{$\tau$}}}
\newcommand{\bfPhi}{\mbox{\boldmath{$\Phi$}}}
% \newcommand{\bfpsi}{\mbox{\boldmath{$\psi$}}}
\newcommand{\bftheta}{\mbox{\boldmath{$\theta$}}}
\newcommand{\bfPsi}{\mbox{\boldmath{$\Psi$}}}
\newcommand{\bfT}{\mbox{\boldmath{${\cal T}$}}}
\newcommand{\bfTheta}{\mbox{\boldmath{$\Theta$}}}
\newcommand{\bfsigma}{\mbox{\boldmath{$\sigma$}}}
\newcommand{\bfvarsigma}{\mbox{\boldmath{$\varsigma$}}}
\newcommand{\bfupsilon}{\mbox{\boldmath{$\upsilon$}}}
\newcommand{\bfgamma}{\mbox{\boldmath{$\gamma$}}}
\newcommand{\bfvarphi}{\mbox{\boldmath{$\varphi$}}}
\newcommand{\bfzeta}{\mbox{\boldmath{$\zeta$}}}
\newcommand{\bfchi}{\mbox{\boldmath{$\chi$}}}
% \newcommand{\bft}{\mbox{\boldmath{$t$}}}
\newcommand{\bfmathcalp}{\mbox{\boldmath{$\mathcal{p}$}}}
\newcommand{\bfLambda}{\mbox{\boldmath{$\Lambda$}}}
\newcommand{\bfzero}{{\bf 0}}
\newcounter{constants}
\setcounter{constants}{0}
\newcommand{\cn}[2]{ \addtocounter{constants}{1} \newcounter{c#1#2}
\setcounter{c#1#2}{\value{constants}} c_{\arabic{c#1#2}} }
\newcommand{\cc}[2]{c_{\arabic{c#1#2}}}
\newcommand{\MBtext}[1]{\marginpar{\red \scriptsize $\leftarrow$ #1}}

%%%%%%%%end new commands%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%odkomentoval jsem, protoze se na tuto analytickou rovnici budu dal v textu odkazovat
When van Genuchten's formula (\ref{vangenuchten-eq}) is substituted in to this integral 
law, the following analytical formula is obtained

  \begin{equation}
  \label{mualem-anal}
  K(h) = \left\{
    \begin{array}{l l}
    K_s \frac{\left( 1- (-\alpha h)^{nm} \left( 1+ (-\alpha h)^n \right)^{-m} \right)^2}{\left(1+(-\alpha h)^n \right)^{\frac{m}{2}}} \\
    \mbox{$\forall$ $h \in$ $(-\infty,0)$};\\
    K_s  \ \mbox{$\forall$ $h \in$ $\langle 0, +\infty)$}\\
    \end{array} \right.
    \end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Since the treatment of the initial condition (\ref{ic}) is nontrivial
in the finite element context, let us say a few words about it. 
At time $t = 0$ the exact solution (\ref{ic}) is constant and it 
does not satisfy the Dirichlet boundary condition on the top edge. 
Therefore we begin at a small positive time $t = 10^{-3}$.
It is accepted that discontinuous initial condition is extremely common
in practice and thus it will appear in the last example (\ref{sec:example-barrier}).

Starting from a coarse low-order mesh, automatic adaptivity is used
to obtain a piecewise-polynomial approximate initial condition 
$h_0$ that approximates the (exact) initial condition $h$ with 
a sufficiently small relative error $TOL$ in the $H^1$-norm. The 
spatial accuracy for all subsequent computations is set to $TOL = 1\%$.
The function $h_0$ is shown in Fig. \ref{fig:init-cond}.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[height=4.5cm]{mesh_init_h1.png}
\end{center}
\vspace{-6mm}
\caption{Initial mesh for $h$-FEM with linear elements (24843 DOF).}
\label{fig:init-meshes-3}
\end{figure}
Implicit Euler method was used for time integration.
In Figs. \ref{fig:init-meshes-1} - \ref{fig:init-meshes-3} 
different colors are used to distinguish between 
different polynomial degrees of finite elements. All elements in these computations 
are quadrilaterals. Notice in Fig. \ref{fig:init-meshes-1} that adaptive $hp$-FEM employs
anisotropic (i.e., directionally different) polynomial degrees in finite 
elements -- this is when an element in the mesh contains two different colors. 
The steady-state solution is shown in Fig. \ref{fig:steady-state}.

@@

It is well known  that
both the van Genuchten's model~(\ref{vangenuchten-eq}) and Mualem's model~(\ref{mualem-anal})
involves a large amount of computer operations for its evaluation (see e.g. Ross~\cite{ross} or Simunek et. al.~\cite{simunek}).
Methods to approximate the unsaturated hydraulic conductivity function were discussed by Miller et al.~\cite{miller}
and Hermite polynomials were recommended to use.

The function $C(h)$ was replaced by Hermit cubic splines (only its first derivative is required 
by the Newton's method) and $K(h)$ was replaced with Hermit quintic spline (its first and second derivative is required 
by the Newton method). The derivatives are obtained by derivating the spline functions.
Via this approximation we cured the problem of singularity of the second derivative 
of $K(h)$,

$$
  \lim_{h \to 0-} K''(h) = -\infty \quad \mbox{but} \quad \lim_{h \to 0+} K''(h) = 0. 
$$ 
The spline approximation $p(h)$ has 
$$
  p''(0) = 0,
$$
as shown in Fig. \ref{fig:kh-approx} (for gravel).
This spline only differs from the original function $K(h)$ in the first interval 
$(-1, 0)$ and in the remaining intervals it is virtually identical to the 
original function. Close to full saturation, the approximating curve's shape is 
very similar to the modified van Genuchten Mualem model that was recommended by Vogel et al. 
in~\cite{vogel}. Compared to the original Vogel's approach, our spline approximation 
preserves the smoothness of the $K(h)$ function in its entire domain of definition.
Unsaturated hydraulic properties of all materials are summarized in Table \ref{tab:mat-example3}.

@@

The previous sections~\ref{sec:compar} -- \ref{sec:example-barrier} presents an application
 of the Richards' equation law on various theoretical and engineer problems. An h-based form~(\ref{rich-final})  numerically approximated by $h$ and $hp$ adaptive finite element method was considered. It is well known (see e.g. Celia, et. al.~\cite{celia}) that improper time discretization and integration techniques of this form of the Richards' equation exhibit in mass balance errors and erroneous estimates of an infiltration depth. 
The error is caused by an improper discrete approximation of the time derivative term $C(h)\frac{\partial h}{\partial t}$, originating from mass (water content) derivative term $\frac{\partial \theta(h)}{\partial t}$.
Higher order time integration methods as considered by Kees et al.~\cite{kees-time-integ} for two-phase flow can treat the mass balance problem. An adaptivity in time as a response to changing conditions is an efficient approach to maintain mass conservation.
Alternatively, in case of a low order time integration method,  Kuraz~et.al.~\cite{kuraz-jcam} recommends time step adaptivity criterion based on retention curve torsion.

The current implementation gives an estimate on error in  spatial approximation, nevertheless the time integration is maintained by 
low order Crank-Nicholson and implicit Euler scheme. The implementation of the Richards equation into the Hermes project is currently  under 
a development and the Crank-Nicholson scheme is not the final solution. More robust method, a so call SDIRKNG (Singly diagonally implicit Runge-Kutta-Nystr\'om  general), see~\cite{sdirk}, is under an evaluation. The time integration will be based on arbitrary Butcher's tables and then the time adaptivity will be enabled by embedded Runge-Kutta's methods by estimating truncation error.

Based on the results obtained from all conducted examples it might be apparent that the $hp$ adaptive method requires a very similar computational CPU demands compared to $h$ adaptivity with $p=2$. The advantage of $hp$ FEM methods appears together with an increasing demands for spatial accuracy due to its exponential convergence. As input data uncertainty is a typical issue for many hydrological problems $h$ adaptivity with $p=2$ can provide a sufficiently good performance for many engineering applications.

@@

Let $\Omega$ be a bounded domain in $\mathbb{R}^2$,
with a Lipschitz boundary $\partial\Omega$. Let
$\Gamma_D$ and $\Gamma_N$ be smooth open disjoint subsets of
$\partial\Omega$ (not necessarily connected) such that
$\partial\Omega = \overline{\Gamma_D}\cup\overline{\Gamma_N}$,
$\textmd{meas}_{n-1}(\partial\Omega\setminus({\Gamma_D}\cup{\Gamma_N}
))=0$ and $\textmd{meas}_{n-1}(\Gamma_D)>0$ and
$\textmd{meas}_{n-1}(\Gamma_N)>0$. In what follows $\bfn$ denotes the outer unit
normal to $\partial\Omega$. For positive $T$ let
$Q_T=\Omega\times[0,T)$.

Let
\begin{displaymath}
{\mathcal{E}}(\overline{\Omega}):=\left\{\bfv\in{\mathcal
C}^\infty(\overline{\Omega})^n; \, {\textmd{supp}\, \bfv}  \cap
\Gamma_D = \emptyset \right\}.
\end{displaymath}
Let the linear space $\mathbb{V}$ be a closure of
${\mathcal{E}}(\overline{\Omega})$ in the norm of $W^{1,2}(\om)$.

The mathematical model of unsaturated flow with initial data and the
mixed boundary conditions is described by the following
initial--boundary value problem
\begin{align}
 C(h)\frac{\partial h}{\partial t}
 - \nabla \cdot \left( K(h) \nabla h \right)
  -  \frac{\partial K(h)}{\partial z}
  &= 0  &&\textmd{in}\; Q_T,
\label{dual_richards}
\\
h(0) &= h^0 &&\textmd{in}\;\Omega, \label{dual_richards_init}
\\
h &= h_D  &&\textmd{on}\;\Gamma_D \times
[0,T),\label{dirichlet_dual}
\\
\frac{\partial h}{\partial \bfn} &= g &&\textmd{on}\;\Gamma_N \times
[0,T). \label{neumann1_dual}
\end{align}

Here we consider all functions to be smooth enough and that the
compatibility condition $h^0=h_D$ on $\Gamma_D$ holds.

The weak formulation of the problem
\eqref{dual_richards}--\eqref{neumann1_dual} is as follows:

Let the given data satisfy
\begin{equation}
h^0 \in W^{1,2}(\Gamma_D),  \; h^0 = h_D(0) \; \textmd{on}\;\Gamma_D,
\end{equation}
\begin{equation}
\begin{split}
g \in \mathcal{C}^1([0,T];L^{\infty}(\Gamma_N)),  \; h_D \in
\mathcal{C}^1([0,T];  \\ L^{\infty}(\Gamma_D)\cap W^{1,2}(\Omega)).
\end{split}
\end{equation}
A function
\begin{equation*}
\begin{split}
h \in \mathcal{C}([0,T);W^{1,2}(\Omega))\cap
\mathcal{C}([0,T); \\ L^{\infty}(\Omega))\cap
\mathcal{C}^1([0,T);L^{2}(\Omega)),
\end{split}
\end{equation*}
such that
$$
h(0)=h_0 \textmd{ in } \Omega, \; h(t)=h_D(t) \textmd{ on } \Gamma_D
\times [0,T),
$$
is a variational solution to the system
\eqref{dual_richards}--\eqref{neumann1_dual} iff
\begin{equation}
\begin{split}
\int_{\Omega}  C(h)\frac{\partial h}{\partial t}\varphi \;{\rm
d}\bfx + \int_{\Omega} K(h)\nabla h \cdot \nabla \varphi \;{\rm
d}\bfx - \\ - \int_{\Omega}\frac{\partial K(h)}{\partial z}\varphi \;{\rm
d}\bfx = \int_{\Gamma_N} g \varphi \;{\rm d}\bfS
\end{split}
\label{discrete-rich}
\end{equation}

holds for all $\varphi \in \mathbb{V}$ and $t\in(0,T)$.

The time discretization is accomplished for the example in section~\ref{sec:example} through a Crank-Nicholson scheme, then in  a compact form the  problem~\eqref{discrete-rich} can be written as

\begin{equation}
\begin{split}
\mathbf{C(h}_n^{}) \frac{\mathbf{h}_{n}-\mathbf{h}_{n-1}}{\Delta t} +  \frac{1}{2}(\mathbf{K(h}_n^{})\mathbf{h}_n + \mathbf{K'(h}_n^{})\mathbf{h}_n  \\ \mathbf{K(h}_{n-1}^{})\mathbf{h}_{n-1} + \mathbf{K'(h}_{n-1}^{})\mathbf{h}_{n-1}) +  \mathbf{f} = 0 
\end{split}
\end{equation}

where subscript $n$ denotes the time step number, $\Delta t$ denotes the time step length and $\mathbf{f}$ denotes the boundary value data.

The examples in sections~\ref{sec:compar} and~\ref{sec:example-barrier} consider time integration method through an implicit Euler scheme, then in a compact form the  problem~\eqref{discrete-rich} can be written as

\begin{equation}
\begin{split}
\mathbf{C(h}_n^{}) \frac{\mathbf{h}_{n}-\mathbf{h}_{n-1}}{\Delta t} +  \mathbf{K(h}_n^{})\mathbf{h}_n + \mathbf{K'(h}_n^{})\mathbf{h}_n +  \mathbf{f} = 0 
\end{split}
\end{equation}

Because of the non-linearity of the Richards equation model both the Picard iteration method and the Newton iteration method was implemented to obtain the solution. Matrix $\mathbf{K'(h)}$ is obtained  from derivatives of $K(h)$ function in respect to $h$.
The Picard iteration procedure is formulated as

\begin{equation*}
\mathbf{C(h}_n^{i-1}) \frac{\mathbf{h}_{n}^i-\mathbf{h}_{n-1}}{\Delta t} +  \mathbf{K(h}_n^{i-1})\mathbf{h}_n + \mathbf{K'(h}_n^{i-1})\mathbf{h}_n^i +  \mathbf{f} = 0 
\end{equation*}

where superscript $i$ denotes the iteration level. The convergency criterion is defined as $||h^i - h^{i-1}|| < \varepsilon$, where $\varepsilon$ is some desired error.

The Newton procedure is formulated as

\begin{equation*}
\begin{split}
\frac{1}{\Delta t} \left( \frac{\partial }{\partial \mathbf{h}} \mathbf{C(h}_n^{i})(\mathbf{h}_n^{i}-\mathbf{h}_{n-1}) + \mathbf{C(h})_{n}^i \right)\Delta \mathbf{h}_n^i + \\ + \left(\frac{\partial }{\partial \mathbf{h}}
\mathbf{K(h)}_n^i\mathbf{h}^i_n + \mathbf{K}(\mathbf{h)}_n^i + \frac{\partial }{\partial \mathbf{h}}
\mathbf{K'(h)}_n^i\mathbf{h}^i_n + \mathbf{K'}(\mathbf{h)}_n^i \right) \Delta \mathbf{h}^i_n = \\ = - \left(\mathbf{C(h_n}^i)\frac{\mathbf{h}_n^i - \mathbf{h}_{n-1}}{\Delta t} + \mathbf{K(h)}_n^i\mathbf{h}^i_n + \mathbf{K'(h)}_n^i\mathbf{h}^i_n \right) ,
\end{split}
\end{equation*}

at the end of each iteration the primary variables are updated as follows

$$\mathbf{h}_n^{i+1} = \mathbf{h}_n^i + \Delta \mathbf{h}^i_n \quad .$$

% The non-linear operator was treated by Picard iteration method and by Newton iteration method. The variational solution 
% to the system \eqref{dual_richards}--\eqref{neumann1_dual} if Picard method considered and implicit Euler time integration considered states that
% 
% \begin{equation}
% \begin{split}
% \frac{1}{\Delta t}(h^{i-1})\frac{\partial h^i}{\partial t}\varphi \;{\rm
% d}\bfx + \int_{\Omega} K(h^{i-1})\nabla h^i \cdot \nabla \varphi \;{\rm
% d}\bfx - \\ - \int_{\Omega}K'(h^{i-1})\frac{\partial h^i}{\partial z}\varphi \;{\rm
% d}\bfx = \int_{\Gamma_N} g \varphi \;{\rm d}\bfS
% \end{split}
% \end{equation}
% 
% where $i$ denotes an iteration level. The convergency criterion is defined as $||h^i - h^{i-1}|| < \varepsilon$, where $\varepsilon$ is some desired error.
% 
% 
% The variational solution 
% to the system \eqref{dual_richards}--\eqref{neumann1_dual} if Newton method considered states that
% 
% \begin{equation}
% \begin{split}
% \int_{\Omega}  C(h^{i-1})\frac{\partial h^i}{\partial t}\varphi \;{\rm
% d}\bfx + \int_{\Omega}  C(h^{i-1})'\frac{\partial h^i}{\partial t}\varphi \;{\rm
% d}\bfx + \\ +  \int_{\Omega} K(h^{i-1})\nabla h^i \cdot \nabla \varphi \;{\rm
% d}\bfx + \\ + \int_{\Omega} K(h^{i-1})'\nabla h^i \cdot \nabla \varphi \;{\rm
% d}\bfx - \\ - \int_{\Omega}K(h^{i-1})'\frac{\partial h^i}{\partial z}\varphi \;{\rm
% d}\bfx - \int_{\Omega}K(h^{i-1})''\frac{\partial h^i}{\partial z}\varphi \;{\rm
% d}\bfx = \\ = \int_{\Gamma_N} g \varphi \;{\rm d}\bfS  \quad .
% \end{split}
% \end{equation}
